### 2014-08-22 Pero_Transcriptome.tex
major update, adding tables

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

@@

Deserts are widely considered one of Earth's harshest environments. Animals living in desert environments are forced to endure intense heat and drought, and in turn, species having evolved in these environments are likely to have evolved specialised mechanisms that may enhance fitness. While living in deserts likely involves a large number of adaptive phenotypes, the ability to osmoregulate -- to maintain the proper water and electrolyte balance -- appears to be paramount \cite{Walsberg:2000uu}. Indeed, the maintenance of water balance in animals is one of the most important physiologic processes for all animals, desert inhabitants or not. Most animals are exquisitely sensitive to changes in osmolality, with slight derangement eliciting physiologic compromise.  When the loss of water exceeds dietary intake, dehydration - and in extreme cases, death - can occur, which suggests that there is strong selection for mechanisms supporting osmoregulation. Understanding these mechanisms will significantly enhance our understanding of the physiologic processes underlying osmoregulation in extreme environments, having implications for studies of human health, conservation, and climate change. \\

The genes and structures responsible for the maintinance of water and electrolyte balance are well characterized in model organisms such as mice \cite{Tatum:2009kp}, rats \cite{Romero:2007fs,Rojek:2006dg,NIELSEN:1995uq}, and humans \cite{Mobasheri:2007tt,Bedford:2003wa,NIELSEN:1999vg}. These studies, many of which have been enabled by newer sequencing technologies, serve as a foundation for studies of renal genomics in non-model organisms. In particular, because researchers have long been interested in desert adaptation, a number of studies have looked at the morphology or expression of single genes in the renal tissues of desert adapted rodents \textit{Phyllotis darwini} \cite{Gallardo:2005fm}, \textit{Psammomys obesus} \cite{Kaissling:1975uu}, and \textit{Perognathus penicillatus} \cite{Altschuler:1979du}. More recently, full renal transcriptomes have been generated for \textit{Dipodomys spectabilis} and \textit{Chaetodipus baileyi} \cite{Marra:2014de} as well as \textit{Abrothrix olivacea} \cite{Giorello:2014iv}. \\

@@

While the elucidation of the mechanisms underlying adaptation to desert survival is beyond the scope of this manuscript, we aim here to lay the groundwork by characterizing the transcriptome from four distinct tissues (brain, liver, kidney, testes). These data will be included in current larger effort aimed at sequencing the entire genome. Further, via sequencing the renal tissue of a total of 15 additional animals, we characterize nucleotide polymorphism and genome wide patterns of natural selection. Together, these investigations will aid in our overarching goal -- to understand the genetic bases of adaptation in \textit{P. eremicus}.

@@

Total RNA was extracted from each tissue using a TRIzol extraction (Invitrogen) following the manufacturers instructions. Because preparation of an RNA library suitable for sequencing is dependent on having high
quality, intact RNA, a small aliquot of each total RNA extract was
analyzed on a Bioanalyzer 2100 (Agilent, Palo Alto, CA, USA).
Following confirmation of sample quality, the reference sequencing libraries
were made using the TruSeq stranded RNA prep kit
(Illumina), while an unstranded TruSeq kit was used to construct the other sequencing libraries. A unique index was ligated to each sample to allow for multiplexed sequencing. Reference libraries (n=4 tissue types) were then pooled to
contain equimolar quantities of each individual library and
submitted for Illumina sequencing using two lanes of 150nt paired end sequencing using the rapid-mode of the HiSeq 2500 sequencer at The Hubbard Center for Genome Sciences (University of New Hampshire). The remaining 14 libraries were similarly multiplexed and sequenced in a mixture of 100nt paired and single end across two lanes of an Illumina HiSeq 2000 at the Vincent G. Coates Gnome Center (University of California, Berkeley).

@@

The raw sequence reads were error corrected using the software bless \cite{Heo:2014cb}, using kmer=25, based on the developers defaultrecommendations. The error corrected adapter and quality trimmed following recommendations from MacManes \cite{MacManes:2014io} and Mbandi \cite{Christoffels:2014bg}. Specifically, adapter sequence contamination was removed, and low quality nucleotides (defined as \textsc{Phred} \textless 2) were removed using the program Trimmomatic version 0.32  \cite{Lohse:2012fg}. Reads from each tissue were assembled using Trinity version released 17 July 2014 \cite{Haas:2013jq}. We used flags indicating the stranded nature of sequencing reads and set maximum allowable physical distance between read pairs to 999nt. The assembly was conducted on a linux workstation with 64 cores and 512Gb RAM. To filter the raw sequence assembly, I downloaded \textit{Mus musculus} cDNA and ncRNA datasets from Ensembl (\url{ftp://ftp.ensembl.org/pub/release-75/fasta/mus_musculus/}), and the \textit{Peromyscus maniculatus} reference transcriptome from NCBI (\url{ftp://ftp.ncbi.nlm.nih.gov/genomes/Peromyscus_maniculatus_bairdii/RNA/}). I used a blastN procedure (default settings, evalue set to 10$^{-10}$) to identify contigs in the \textit{P. eremicus} dataset that are likely biological in origin. This procedure, when a reference dataset is available, retains more putative transcripts that a strategy employing expression-based filtering (remove if TMP \textless 1) of the raw assembly. I then concatenated the filtered assemblies from each tissue into a single file, reducing redundancy using the software cd-hit-est \cite{Li:2006hr} using default setting except that sequences were clustered based on 95\% sequence similarity. The resulting assembly file was characterized using the software package transrate (\url{https://github.com/Blahah/transrate}). \\

@@

To identify patterns of gene expression unique to each tissue type, I mapped sequence reads from each tissue type to the reference assembly using bwa-mem. We estimated expression individually for the four tissues using default settings of the software eXpress \cite{Roberts:2012dh}. Interesting patterns of expression, including instances where expression was limited to a single tissue type were identified and visualized. \\

@@

I then estimated gene expression on each of these tissue-specific datasets, which allowed me to understand expression patterns in the multiple tissues ({\hypertarget{Figure 1}{Figure 1}}). After expression estimation, the filtered assemblies were concatenated together, and after removal of redundancy with cd-hit-est, 123,123 putative transcripts remained. From this filtered concatenated dataset, I extracted 71626 putative coding sequences (72Mb). Of these 71626 sequences, 38221 were complete exons (containing both start and stop codons), while other were either truncated at the 5-prime end (20239 sequences), 3-prime end (6445 sequences), or were internal (6721 sequencing having neither stop nor start codon).

@@

@@

%\section*{References}
% The bibtex filename
\bibliography{formatted.bib}

>>>>

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}
\usepackage{chngpage}
% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

@@

Deserts are widely considered one of Earth's most harsh environments. Animals living in desert environments are forced to endure intense heat and drought, and as a result, species having evolved in these environments are likely to posses specialized mechanisms that may enhance fitness. While living in deserts likely involves a large number of adaptive traits, the ability to osmoregulate -- to maintain the proper water and electrolyte balance -- appears to be paramount \cite{Walsberg:2000uu}. Indeed, the maintenance of water balance in animals is one of the most important physiologic processes for all organisms, whether they be desert inhabitants or not. Most animals are exquisitely sensitive to changes in osmolality, with slight derangement eliciting physiologic compromise.  When the loss of water exceeds dietary intake, dehydration - and in extreme cases, death - can occur. This process suggests that there is strong selection for mechanisms supporting osmoregulation. Understanding these mechanisms will significantly enhance our understanding of the physiologic processes underlying osmoregulation in extreme environments, having implications for studies of human health, conservation, and climate change. \\

The genes and structures responsible for the maintenance of water and electrolyte balance are well characterized in model organisms such as mice \cite{Tatum:2009kp}, rats \cite{Romero:2007fs,Rojek:2006dg,NIELSEN:1995uq}, and humans \cite{Mobasheri:2007tt,Bedford:2003wa,NIELSEN:1999vg}. These studies, many of which have been enabled by newer sequencing technologies, serve as a foundation for studies of renal genomics in non-model organisms. In particular, because researchers have long been interested in desert adaptation, a number of studies have looked at the morphology or expression of single genes in the renal tissues of desert adapted rodents \textit{Phyllotis darwini} \cite{Gallardo:2005fm}, \textit{Psammomys obesus} \cite{Kaissling:1975uu}, and \textit{Perognathus penicillatus} \cite{Altschuler:1979du}. More recently, full renal transcriptomes have been generated for \textit{Dipodomys spectabilis} and \textit{Chaetodipus baileyi} \cite{Marra:2014de} as well as \textit{Abrothrix olivacea} \cite{Giorello:2014iv}. \\

@@

While the elucidation of the mechanisms underlying adaptation to desert survival is beyond the scope of this manuscript, we aim here to lay the groundwork by characterizing the transcriptome from four distinct tissues (brain, liver, kidney, testes). These data will be included in current larger effort aimed at sequencing the entire genome. Further, via sequencing the renal tissue of a total of 15 additional animals, we characterize nucleotide polymorphism and genome wide patterns of natural selection. Together, these investigations will aid in our overarching goal -- to understand the genetic bases of adaptation to deserts in \textit{P. eremicus}.

@@

Total RNA was extracted from each tissue using a TRIzol extraction (Invitrogen) following the manufacturers instructions. Because preparation of an RNA library suitable for sequencing is dependent on having high
quality, intact RNA, a small aliquot of each total RNA extract was
analyzed on a Bioanalyzer 2100 (Agilent, Palo Alto, CA, USA).
Following confirmation of sample quality, the reference sequencing libraries
were made using the TruSeq stranded RNA prep kit
(Illumina), while an unstranded TruSeq kit was used to construct the other sequencing libraries. A unique index was ligated to each sample to allow for multiplexed sequencing. Reference libraries (n=4 tissue types) were then pooled to
contain equimolar quantities of each individual library and
submitted for Illumina sequencing using two lanes of 150nt paired end sequencing using the rapid-mode of the HiSeq 2500 sequencer at The Hubbard Center for Genome Sciences (University of New Hampshire). The remaining 15 libraries were similarly multiplexed and sequenced in a mixture of 100nt paired and single end across two lanes of an Illumina HiSeq 2000 at the Vincent G. Coates Gnome Center (University of California, Berkeley).

@@

The raw sequence reads were error corrected using the software bless \cite{Heo:2014cb}, using kmer=25, based on the developers default recommendations. The error corrected adapter and quality trimmed following recommendations from MacManes \cite{MacManes:2014io} and Mbandi \cite{Christoffels:2014bg}. Specifically, adapter sequence contamination was removed, and low quality nucleotides (defined as \textsc{Phred} \textless 2) were removed using the program Trimmomatic version 0.32  \cite{Lohse:2012fg}. Reads from each tissue were assembled using Trinity version released 17 July 2014 \cite{Haas:2013jq}. We used flags indicating the stranded nature of sequencing reads and set maximum allowable physical distance between read pairs to 999nt. The assembly was conducted on a linux workstation with 64 cores and 512Gb RAM. To filter the raw sequence assembly, I downloaded \textit{Mus musculus} cDNA and ncRNA datasets from Ensembl (\url{ftp://ftp.ensembl.org/pub/release-75/fasta/mus_musculus/}), and the \textit{Peromyscus maniculatus} reference transcriptome from NCBI (\url{ftp://ftp.ncbi.nlm.nih.gov/genomes/Peromyscus_maniculatus_bairdii/RNA/}). I used a blastN procedure (default settings, evalue set to 10$^{-10}$) to identify contigs in the \textit{P. eremicus} dataset that are likely biological in origin. This procedure, when a reference dataset is available, retains more putative transcripts that a strategy employing expression-based filtering (remove if TMP \textless 1) of the raw assembly. I then concatenated the filtered assemblies from each tissue into a single file, reducing redundancy using the software cd-hit-est \cite{Li:2006hr} using default setting except that sequences were clustered based on 95\% sequence similarity. %The resulting assembly file was characterized using the software package transrate (\url{https://github.com/Blahah/transrate}). \\

@@

To identify patterns of gene expression unique to each tissue type, I mapped sequence reads from each tissue type to the reference assembly using bwa-mem \cite{Li:2013wn}. We estimated expression individually for the four tissues using default settings of the software eXpress \cite{Roberts:2012dh}. Interesting patterns of expression, including instances where expression was limited to a single tissue type were identified and visualized. \\

@@

I then estimated gene expression on each of these tissue-specific datasets, which allowed me to understand expression patterns in the multiple tissues ({\hypertarget{Figure 1}{Figure 1}}). After expression estimation, the filtered assemblies were concatenated together, and after removal of redundancy with cd-hit-est, 123,123 putative transcripts remained. From this filtered concatenated dataset, I extracted 71626 putative coding sequences (72Mb). Of these 71626 sequences, 38221 were complete exons (containing both start and stop codons), while other were either truncated at the 5-prime end (20239 sequences), 3-prime end (6445 sequences), or were internal (6721 sequencing having neither stop nor start codon). \\

\begin{adjustwidth}{-.35in}{-.05in}% adjust the L and R margins by 1 inch
\begin{tabular}{ l l l l l l l}
Transcript ID	&	Testes	&	Liver	&	Kidney	&	Brain	&	Genbank ID	&	Gene ID	\\
Transcript\_83842	&	2.05E+03	&	6.40E+03	&	1.03E+04	&	5.47E+03	&	DQ073446.1	&	COX2	\\
Transcript\_126459	&	1.43E+01	&	2.22E+04	&	2.77E+01	&	6.73E+00	&	XM\_006991665.1	&	Alb	\\
Transcript\_128937	&	4.39E+00	&	1.91E+04	&	4.74E+02	&	2.23E+00	&	XM\_007627625.1	&	Apoa2	\\
Transcript\_81233	&	1.71E+03	&	5.23E+03	&	6.11E+03	&	3.08E+03	&	XM\_006993867.1	&	Fth1	\\
Transcript\_94125	&	3.67E+01	&	1.08E+04	&	2.09E+03	&	2.75E+00	&	XM\_006977178.1	&	CytP450	\\
Transcript\_119945	&	5.03E+03	&	1.15E+03	&	1.33E+03	&	3.71E+03	&	XM\_008686011.1	&	Ubb	\\
Transcript\_5977	&	4.95E+00	&	1.01E+04	&	3.05E+02	&	3.58E+02	&	XM\_006978668.1	&	Tf	\\
Transcript\_4057	&	2.62E+01	&	9.32E+03	&	1.34E+02	&	8.38E+01	&	XM\_006994871.1	&	Apoc1	\\
Transcript\_112523	&	4.07E+02	&	7.36E+03	&	7.78E+02	&	9.54E+02	&	XM\_006994872.1	&	Apoe	\\
Transcript\_98376	&	1.98E+00	&	8.66E+03	&	1.02E+00	&	2.68E+00	&	XM\_006970208.1	&	Ttr	\\ \end{tabular}
\begin{quote}
\small{Table 2. The 10 transcripts with the highest mean TPM (transcripts per million).}
\end{quote}  
\end{adjustwidth}

@@

\begin{adjustwidth}{-.75in}{-.75in}% adjust the L and R margins by 1 inch
\begin{tabular}{ l l l l }
\textbf{Transcript ID} & \textbf{GenBank ID} & \textbf{Description} & \textbf{Tajima's D}\\
Transcript\_49049 & XM\_006533884.1 & heterogeneous nuclear ribonucleoprotein H1 (Hnrnph1) & 3.26\\
Transcript\_38378 & XM\_006522973.1 & Son DNA binding protein (Son) & 3.19\\
Transcript\_126187 & NM\_133739.2 & transmembrane protein 123 (Tmem123) & 3.02\\
Transcript\_70953 & XM\_006539066.1 & chloride channel Kb (Clcnkb) & 2.96 \\
Transcript\_37736 & XM\_006997718.1 & h-2 class I histocompatibility antigen & 2.92 \\
Transcript\_21448 & XM\_006986148.1 & zinc finger protein 624-like & 2.84\\
Transcript\_47450 & NM\_009560.2 zinc & finger protein 60 (Zfp60) & 2.82\\
Transcript\_122250 & XM\_006539068.1 & chloride channel Kb (Clcnkb) & 2.81\\
Transcript\_78367 & XM\_006496814.1 & CDC42 binding protein kinase alpha (Cdc42bpa) & 2.78 \\
Transcript\_96470 & XM\_006987129.1 & interferon-inducible GTPase 1-like & 2.77 \\
 \end{tabular}
\begin{quote}
\small{Table 4. The 10 transcripts with the highest values for Tajima's D, which is suggestive of positive selection.}
\end{quote}  
\end{adjustwidth}

\begin{adjustwidth}{-.75in}{-.75in}% adjust the L and R margins by 1 inch
\begin{tabular}{ l l l l }
\textbf{Transcript ID} & \textbf{GenBank ID} & \textbf{Description} & \textbf{Tajima's D} \\
Transcript\_84359 & XM\_006991127.1 & nuclear receptor coactivator 3 (Ncoa3) & -2.82\\
Transcript\_87121 & XM\_006970128.1 & methyl-CpG binding domain protein 2 (Mbd2) & -2.82 \\
Transcript\_125755 & EU053203.1 & alpha globin gene cluster & -2.78\\
Transcript\_87128 & XM\_006976644.1 & membrane-associated ring finger (March5) & -2.76 \\
Transcript\_55468 & XM\_006978377.1 & Vpr (HIV-1) binding protein (Vprbp) & -2.75\\
Transcript\_116042 & XM\_006980811.1 & membrane associated guanylate kinase (Magi3) & -2.75  \\
Transcript\_18966 & XM\_006982814.1 & ubiquitin protein ligase E3 component n-recognin 5 (Ubr5) & -2.75 \\
Transcript\_122204 & XM\_008772511.1 & zinc finger protein 612 (Zfp612) & -2.75 \\
Transcript\_100550 & XM\_006971297.1 & bromodomain adjacent to zinc finger domain, 1B (Baz1b) & -2.74\\
Transcript\_33267 & XM\_006975561.1 & pumilio RNA-binding family member 1 (Pum1) & -2.75\\
 \end{tabular}
\begin{quote}
\small{Table 3. The 10 transcripts with the lowest values for Tajima's D, which is suggestive of purifying selection.}
\end{quote}  
\end{adjustwidth}

@@

%\section*{References}
% The bibtex filename
\bibliography{/Users/macmanes/Documents/pero_transcriptome/biblio.bib}
